---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# iSEEpathways

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/iSEE/iSEEpathways)](https://github.com/iSEE/iSEEpathways/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/iSEE/iSEEpathways)](https://github.com/iSEE/iSEEpathways/pulls)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check-bioc](https://github.com/iSEE/iSEEpathways/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/iSEE/iSEEpathways/actions)
[![Codecov test coverage](https://codecov.io/gh/iSEE/iSEEpathways/branch/main/graph/badge.svg)](https://app.codecov.io/gh/iSEE/iSEEpathways?branch=main)
<!-- badges: end -->

The goal of `iSEEpathways` is to provide panels to facilitate the interactive visualisation of pathway analysis results in `r BiocStyle::Biocpkg("iSEE")` applications.

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `iSEEpathways` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("iSEEpathways")
```

And the development version from [GitHub](https://github.com/iSEE/iSEEpathways) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("iSEE/iSEEpathways")
```

## Example

This is a basic example which shows you how to load the package:

```{r example, message=FALSE, warning=FALSE, eval = requireNamespace('iSEEpathways')}
library("iSEEpathways")
library("org.Hs.eg.db")
library("fgsea")
library("iSEE")

# Example data ----

## Pathways
pathways <- select(org.Hs.eg.db, keys(org.Hs.eg.db, "SYMBOL"), c("GOALL"), keytype = "SYMBOL")
pathways <- subset(pathways, ONTOLOGYALL == "BP")
pathways <- unique(pathways[, c("SYMBOL", "GOALL")])
pathways <- split(pathways$SYMBOL, pathways$GOALL)
len_pathways <- lengths(pathways)
pathways <- pathways[len_pathways > 15 & len_pathways < 500]

## Features
set.seed(1)
# simulate a score for all genes found across all pathways
feature_stats <- rnorm(length(unique(unlist(pathways))))
names(feature_stats) <- unique(unlist(pathways))
# arbitrarily select a pathway to simulate enrichment
pathway_id <- "GO:0046324"
pathway_genes <- pathways[[pathway_id]]
# increase score of genes in the selected pathway to simulate enrichment
feature_stats[pathway_genes] <- feature_stats[pathway_genes] + 1

# fgsea ----

set.seed(42)
fgseaRes <- fgsea(pathways = pathways, 
                  stats    = feature_stats,
                  minSize  = 15,
                  maxSize  = 500)
head(fgseaRes[order(pval), ])

# iSEE ---

ngenes <- length(feature_stats)
cnts <- matrix(rnbinom(n=ngenes*2, mu=100, size=1/0.5), nrow=ngenes)
rownames(cnts) <- names(feature_stats)
se <- SummarizedExperiment(assay = list(counts = cnts))

se <- embedPathwaysResults(fgseaRes, se, name = "fgsea", class = "fgsea", pathwayType = "GO",
                           pathways = pathways, stats = feature_stats)

app <- iSEE(se, initial = list(
  PathwaysTable(ResultName="fgsea", PanelWidth = 6L),
  FgseaEnrichmentPlot(ResultName="fgsea", PathwayId="GO:0000002", PanelWidth = 6L)
))

if (interactive()) {
  shiny::runApp(app)
}
```

## Citation

Below is the citation output from using `citation('iSEEpathways')` in R. Please
run this yourself to check for any updates on how to cite __iSEEpathways__.

```{r 'citation', eval = requireNamespace('iSEEpathways')}
print(citation('iSEEpathways'), bibtex = TRUE)
```

Please note that the `iSEEpathways` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `iSEEpathways` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://iSEE.github.io/iSEEpathways) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.

## Code of Conduct

Please note that the iSEEpathways project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.
