---
title: "Working with the Gene Ontology"
author: 
  - name: Kevin Rue-Albrecht
    affiliation:
    - University of Oxford
    email: kevin.rue-albrecht@imm.ox.ac.uk
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('iSEEpathways')`"
vignette: >
  %\VignetteIndexEntry{Working with the Gene Ontology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

```{r, eval=!exists("SCREENSHOT"), include=FALSE}
SCREENSHOT <- function(x, ...) knitr::include_graphics(x)
```

```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    iSEEpathways = citation("iSEEpathways")[1]
)
```

# Example data

Generate pathway analysis results using `r BiocStyle::Biocpkg("fgsea")`.

```{r "start", message=FALSE, warning=FALSE}
library("org.Hs.eg.db")
library("fgsea")

# Example data ----

## Pathways
pathways <- select(org.Hs.eg.db, keys(org.Hs.eg.db, "SYMBOL"), c("GOALL"), keytype = "SYMBOL")
pathways <- subset(pathways, ONTOLOGYALL == "BP")
pathways <- unique(pathways[, c("SYMBOL", "GOALL")])
pathways <- split(pathways$SYMBOL, pathways$GOALL)
len_pathways <- lengths(pathways)
pathways <- pathways[len_pathways > 15 & len_pathways < 500]

## Features
set.seed(1)
# simulate a score for all genes found across all pathways
feature_stats <- rnorm(length(unique(unlist(pathways))))
names(feature_stats) <- unique(unlist(pathways))
# arbitrarily select a pathway to simulate enrichment
pathway_id <- "GO:0046324"
pathway_genes <- pathways[[pathway_id]]
# increase score of genes in the selected pathway to simulate enrichment
feature_stats[pathway_genes] <- feature_stats[pathway_genes] + 1

# fgsea ----

set.seed(42)
fgseaRes <- fgsea(pathways = pathways, 
                  stats    = feature_stats,
                  minSize  = 15,
                  maxSize  = 500)
head(fgseaRes[order(pval), ])
```

Embed the `r BiocStyle::Biocpkg("fgsea")` results in the `r BiocStyle::Biocpkg("SummarizedExperiment")` object.

```{r, message=FALSE, warning=FALSE}
library("SummarizedExperiment")
library("iSEEpathways")
ngenes <- length(feature_stats)
cnts <- matrix(rnbinom(n=ngenes*2, mu=100, size=1/0.5), nrow=ngenes)
rownames(cnts) <- names(feature_stats)
se <- SummarizedExperiment(assay = list(counts = cnts))

fgseaRes <- as.data.frame(fgseaRes)
rownames(fgseaRes) <- fgseaRes$pathway
metadata(se) <- list("iSEEpathways" = list("fgsea" = fgseaRes))
```

## Live app

Configure the app option `PathwaysTable.select.details` to define a function that display information about the currently selected GO term.

```{r, message=FALSE, warning=FALSE}
library("iSEE")
library("GO.db")
library("shiny")
go_details <- function(x) {
    info <- select(GO.db, x, c("TERM", "ONTOLOGY", "DEFINITION"), "GOID")
    html <- list(p(strong(info$GOID), ":", info$TERM, paste0("(", info$ONTOLOGY, ")")))
    if (!is.na(info$DEFINITION)) {
        html <- append(html, list(p(info$DEFINITION)))
    }
    tagList(html)
}
se <- registerAppOptions(se, PathwaysTable.select.details = go_details)
```

Configure the app initial state and launch the live app.


```{r, message=FALSE}
app <- iSEE(se, initial = list(
  PathwaysTable(ResultName="fgsea", PanelWidth = 12L)
))

if (interactive()) {
  shiny::runApp(app)
}
```

```{r, echo=FALSE, out.width="100%"}
SCREENSHOT("screenshots/gene_ontology.png", delay=20)
```


# Reproducibility

The `r Biocpkg("iSEEpathways")` package `r Citep(bib[["iSEEpathways"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("gene-ontology.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("gene-ontology.Rmd", tangle = TRUE)
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
